---
title: "Exercise 05"
---

# Play with Geospatial Data {.unnumbered}

## Learning Objectives {.unnumbered}

- Introduce spatial data visualization in ***R*** using **shapefiles**, **gpx** files and the **simple features** standard
- These can be the basis for lots of common visualizations, including drawing polygon data (e.g., countries, bodies of water), lines (e.g., roads, utility lines, animal travel routes), and points (e.g., cities, observation locations) and attaching aesthetics to them (e.g., size, shape, color) to create an analytical insight

#### Preliminaries {.unnumbered}

-   Install and load in the following libraries: {rgdal}, {sp}, {sf}, {tmap}, {adehabitatHR}, {ggspatial}, {cowplot}, {ggsn}, {leaflet}

```{r}
library(rgdal) # for readOGR()
library(sp) # for spatial points (sp) spatial data
library(sf) # for simple feature (sf) spatial data
library(tmap) # for fast plotting of thematic maps
library(adehabitatHR) # for MCP and KDE home ranges
library(ggspatial) # for annotating maps
library(cowplot) # for assembling pretty plots
library(ggsn) # for pretty N arrow
```

Thinking about spatial data the way ***R*** does:

Many GIS programs (e.g., ArcGIS, QGIS) make it easy for users to create maps by loading a specially formated file with geographic information identifiers and data to plot. These programs do a lot of work behind the scenes to translate file into a geospatial shape.

In ***R***, we have to do this work ourselves. ***R*** has no innate knowledge of what we want to plot; we typicall need to every detail. The manual nature of dealing with geospatial data in ***R*** can cause some headaches, as we often need to hunt down all of the information in order to provide it to ***R*** to visualize. Once we have the information, however, ***R***’s graphics routines allows us lots of flexibility. We can add lots of different kinds of information as individual layers to the same plot and style them as we please.

In this exercise, we will import and plot several formats of geospatial data (shapefiles, GPX files, .csv files) using data from Ecuador as an example.

## Important Vocabulary:

**Shapefile** is a data storage format where you can store different types of geographic information and geospatial data such as location, shape, line and points and their various details such as names etc. The information in a shapefile is actually a big set of files with large data in them combined within one shapefile. Many GIS applications and softwares such as ArcGIS use shp file for their visualizations, mapping and analyses. But since they are not usually free it’s a great feature that R has as a free tool.

**GPX files** is a another data storage format...

**Point data** represent distinct and separate points and are commonly defined by their geographic coordination and can be located within a polygon, for example location of a bridge and an archaeological site can be shown as a point.


The `readOGR()` function can be used to read in ESRI shapefiles or ".gpx" files. Here, we load in data in various formats for my research site in Ecuador...

```{r plotting study location}
countries <- readOGR(dsn=paste0("data/",spatial_dir),layer="SA-countries")
class(countries) # SpatialPolygonsDataFrame
ecuador <- readOGR(dsn=paste0("data/",spatial_dir),layer="ecuador")
reserves <- readOGR(dsn=paste0("data/",spatial_dir),layer="reserves-ecuador")
site <- readOGR(dsn=paste0("data/",spatial_dir),layer="tbs-site")
class(site) # SpatialPointsDataFrame
riverfile <- "smooth-rio-tiputini.gpx"
river <- readOGR(dsn=paste0("data/",spatial_dir,"/",riverfile), layer = "routes")
class(tiputini) #SpatialLinesDataFrame"
trailsfile <- "trails.gpx"
trails <- readOGR(dsn=paste0("data/",spatial_dir,"/",trailsfile), layer = "routes")
```

The {sf} package is used to convert to **simple features** format.

```{r}
countries <- st_as_sf(countries)
ecuador <- st_as_sf(ecuador)
reserves <- st_as_sf(reserves)
site <- st_as_sf(site)
river <- st_as_sf(river)
trails <- st_as_sf(trails)
```

{ggplot} can be used for mapping **simple features** with `geom_sf()`

```{r}
sa <- ggplot() +
  geom_sf(data = countries, aes(), fill=NA, color="black", size = 0.2) +
  theme_void() +
  theme(legend.position = "none")

sa <- sa +
  geom_sf(data = ecuador, aes(), alpha=0.5, fill="thistle3", color="black", size = 0.2)

sa <- sa +
  geom_sf(data = site, aes(), color="black", fill="red", size=2, shape=23)

sa <- sa +
  annotation_north_arrow(
    location = "br",
     which_north = "true",
     pad_x = unit(0.75, "in"),
     pad_y = unit(0.5, "in"),
     style = north_arrow_fancy_orienteering)

sa # pretty plot of South America with Ecuador highlighted

e <- ggplot() +
  geom_sf(data = ecuador, aes(), fill="thistle3", alpha=0.5, color="black", size = 0.5) +
  geom_sf(data = reserves, aes(), fill="dark green", alpha=0.5, color=NA, size = 0.1) +
  scale_y_continuous(
    breaks=seq(-5, 1, by=1), limits = c(-5.5, 1.5)) +
  scale_x_continuous(
    breaks=seq(-81, -75, by=1), limits = c(-81.5, -74.5)) +
	theme_light() +
	labs(x="Longitude", y="Latitude")

e <- e +
  geom_sf(data = site, aes(), color="black", fill="red", alpha=0.8, size=3, shape=23)

e <- e +
  annotation_scale(
    location = "br", width_hint = 0.25,
    pad_x = unit(0.5, "in"),
    pad_y = unit(0.5, "in")) +
  annotation_north_arrow(
    location = "br", which_north = "true",
    pad_x = unit(0.75, "in"), pad_y = unit(0.8, "in"),
    style = north_arrow_fancy_orienteering)

e # pretty plot of Ecuador

tbs <- ggplot() +
  geom_sf(data=trails, aes(color = "Trails"), size = 0.5) +
  geom_sf(data=river, aes(color = "Rio Tiputini"), size = 1.5) +
  annotation_scale(
    location = "bl", width_hint = 0.5) +
  annotation_north_arrow(
    location = "bl", which_north = "true",
    pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"),
    style = north_arrow_fancy_orienteering) +
  scale_y_continuous(
    breaks=seq(-0.65, -0.62, by=0.01), limits = c(-0.615, -0.655)) +
  scale_x_continuous(
    breaks=seq(-76.17, -76.13, by=0.01), limits = c(-76.175, -76.125)) +
  theme_light() +
  labs(x="Longitude", y="Latitude") +
  scale_color_manual(
    values = c("Trails" = "grey", "Rio Tiputini" = "blue")) +
  theme(legend.position="none")
tbs # pretty plot of TBS
```

```{r}
library(tidyverse)
library(lubridate)
extract <- read_csv(paste0("data/",spatial_dir,"/","locations.csv"))
coords <- extract %>% dplyr::select(c("mean_longitude", "mean_latitude")) # coordinates as latlon
data <- extract %>%
  mutate(date = date(point)) %>%
  dplyr::select(date, mean_ltime)
crs <- CRS("+init=epsg:4326") # proj4string of coords
locs <- SpatialPointsDataFrame(coords = coords, data = data, proj4string = crs)
class(locs)
locs <- spTransform(locs, CRS("+proj=utm +south +zone=18 +datum=WGS84 +ellps=WGS84 +units=m +no_defs"))
class(locs)
locsSF <- st_as_sf(locs)
class(locsSF)
p <- ggplot() + geom_sf(data = locsSF)
p
p <- tbs + geom_sf(data = locsSF)
p
```
